Towards Practical First-Order Model Counting

### 2024-07-11 doc/paper/paper.tex
AAAI Author Kit

\documentclass{article}
\pdfpagewidth=8.5in
\pdfpageheight=11in

\usepackage{kr}

\usepackage{times}
\usepackage{soul}
\usepackage{url}
\usepackage[hidelinks]{hyperref}
\usepackage[utf8]{inputenc}
\usepackage[small]{caption}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{booktabs}
\urlstyle{same}

\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage[capitalize,noabbrev]{cleveref}
\usepackage{microtype}
\usepackage{mathtools}
\usepackage{forest}
\usepackage{siunitx}
\usepackage{amsfonts}
\usepackage{pifont}
\usepackage[inline]{enumitem}

@@

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{fact}{Observation}
\newtheorem{corollary}{Corollary}
\newtheorem{assumption}{Assumption}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}

\pdfinfo{
/TemplateVersion (KR.2022.0, KR.2023.0, KR.2024.0)
}

@@

\newcommand{\crefrangeconjunction}{--}

\crefname{line}{line}{lines}

\crefname{fact}{Observation}{Observations}
\crefname{assumption}{Assumption}{Assumptions}

\crefalias{enumi}{type}
\crefname{type}{Type}{Types}
\creflabelformat{type}{#2\textup{#1}#3}

\crefalias{enumi}{step}
\crefname{step}{Step}{Steps}
\creflabelformat{step}{#2\textup{#1}#3}

\crefalias{clause}{equation}
\crefname{clause}{Clause}{Clauses}
\creflabelformat{clause}{#2\textup{(#1)}#3}

\crefalias{formula}{equation}
\crefname{formula}{Formula}{Formulas}
\creflabelformat{formula}{#2\textup{(#1)}#3}

\DeclareMathOperator{\CR}{CR}
\DeclareMathOperator{\DR}{DR}
\DeclareMathOperator{\Reff}{Ref}

\DeclareMathOperator{\Doms}{Doms}

\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}

\newcommand{\expr}{\mathtt{expr}}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\FO}{$\mathsf{FO}$}
\newcommand{\UFO}{$\mathsf{UFO}^{2} + \mathsf{EQ}$}

\newcommand{\Cranetwo}{\textsc{Crane2}}
\newcommand{\Cranebfs}{\textsc{Crane2-BFS}}
\newcommand{\Cranegreedy}{\textsc{Crane2-Greedy}}

@@

\title{Towards Practical First-Order Model Counting}

\author{%
  Ananth K. Kidambi$^1$\footnote{The first and second authors contributed equally and were affiliated with the National University of Singapore during the completion of this work.}\and
Guramrit Singh$^1$\and
Paulius Dilkas$^2$\and
Kuldeep S. Meel$^3$ \\
\affiliations
$^1$Indian Institute of Technology Bombay, Mumbai, India\\
$^2$National University of Singapore, Singapore, Singapore\\
$^3$University of Toronto, Toronto, Canada\\
\emails
210051002@iitb.ac.in,
guramrit@iitb.ac.in,
paulius.dilkas@nus.edu.sg,
meel@cs.toronto.edu
}% TODO: update affiliation and email

@@

The complexity of WFOMC is typically characterised in terms of \emph{data
  complexity}. This involves fixing the formula and determining whether an
algorithm exists that can compute the WFOMC in time polynomial with respect to
the domain size(s). If such an algorithm exists, the formula is called
\emph{liftable}~\cite{DBLP:conf/starai/JaegerB12}. Beame et
al.~\shortcite{DBLP:conf/pods/BeameBGS15} demonstrated the existence of an
unliftable formula with three variables. It is also known that all formulas with
up to two variables are
liftable~\cite{DBLP:conf/nips/Broeck11,DBLP:conf/kr/BroeckMD14}. The liftable
fragment of formulas with two variables has been expanded with various
axioms~\cite{DBLP:conf/aaai/TothK23,DBLP:journals/ai/BremenK23}, counting
quantifiers~\cite{DBLP:journals/jair/Kuzelka21} and in other
ways~\cite{DBLP:conf/nips/KazemiKBP16}.

@@

  % TODO: Yes, as mentioned at the end of Section 2.1, clauses c and d written
  % on separate lines are to be interpreted as the conjunction of c and d (maybe
  % move that sentence here?).
  Equivalently, in \FO{} we would write
  \begin{equation}\label[formula]{eq:fo}
    \begin{gathered}
      \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y)\\
      \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z.
    \end{gathered}
  \end{equation}
  The first clause asserts that each $x$ must have at least one corresponding
  $y$, while the second statement adds the condition that if $x$ is mapped to
  both $y$ and $z$, then $y$ must equal $z$. It is important to note that
  \cref{eq:fo} is written with two domains instead of just one. However, we can
  still determine the number of endofunctions by assuming that the sizes of
  $\Gamma$ and $\Delta$ are equal. This formulation, as observed by Dilkas and
  Belle~\shortcite{DBLP:conf/kr/DilkasB23}, can prove beneficial in enabling
  first-order knowledge compilation algorithms to find efficient solutions.
\end{example}

@@

% TODO: F missing (also not explained that it’s not needed below). E is not explained other than that we take the example from X. Would not be a problem if it is motivated differently and the focus is on post-processing (defining inputs properly) and not on the entire algorithm.
% TODO: include a high-level overview of first-order knowledge compilation and the meaning of the resulting functions. For the most part, the compilation process that constructs these function definitions can be treated as a black box for the purposes of this work. However, a few high-level details still ought to be mentioned:
% 1. Each algebraic variable ("m", "n", and "l" in the example) corresponds to a domain size.
% 2. \mathcal{D} indicates that "m" denotes the size of \Gamma, "n" denotes the size of \Delta, etc.
% 3. \mathcal{E} can contain any (positive) number of functions.
% 4. Function denoted by "f" represents the solution to the model counting problem.
% 3. The model count of the original formula is computed by evaluating f on the desired domain sizes. For example, f(3, 2) is the model count when |\Gamma| = 3, and |\Delta| = 2.
\begin{example}[Counting bijections]\label{example:overall}
  Consider the following formula (previously examined by Dilkas and
  Belle~\shortcite{DBLP:conf/kr/DilkasB23}) that defines predicate $P$ as a
  bijection between two sets $\Gamma$ and $\Delta$:
  \[
    \begin{gathered}
      \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y)\\
      \forall y \in \Delta\text{. }\exists x \in \Gamma\text{. }P(x, y)\\
      \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z\\
      \forall x, z \in \Gamma\text{. }\forall y \in \Delta\text{. }P(x, y) \land P(z, y) \Rightarrow x = z.
    \end{gathered}
  \]
  We specifically examine the first solution returned by \Cranetwo{} for this
  formula.

@@

\emph{Smoothness} is a property originating in propositional knowledge
compilation, where it refers to the situation where all disjuncts in a
disjunction node contain the same atoms~\cite{darwiche2001tractable}. Van den
Broeck et al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11} extend this concept to
first-order logic, introducing set-disjunction and inclusion-exclusion nodes in
addition to disjunction.

@@

The remainder of this section presents an extension to the smoothing algorithm
of Van den Broeck et al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11}.
\Cref{sec:smoothingbase} explains the role of smoothing in the base-case-finding
algorithm described in \cref{sec:main}. \Cref{sec:smoothingfcg1} demonstrates
how to adapt smoothing to the compilation rules introduced by Dilkas and
Belle~\shortcite{DBLP:conf/kr/DilkasB23}.

@@

Before discussing the proposed changes to smoothing, let us briefly review the
compilation rules and their corresponding node types introduced by Dilkas and
Belle~\shortcite{DBLP:conf/kr/DilkasB23}. \emph{Domain recursion} involves
selecting a domain $\Delta$, introducing a new constant $c \in \Delta$, and
modifying the formula based on two possibilities for each variable $x$
quantified over $\Delta$: $x = c$ or $x \ne c$. The resulting node is denoted as
$\DR(c \in \Delta)$. \emph{Constraint removal} applies to formulas where each
variable $x$ quantified over a domain $\Delta$ is followed by the inequality
constraint $x \ne c$. These formulas can be rewritten so that each clause begins
with $\forall x \in \Delta\text{. }x \ne c \Rightarrow \ldots$ Constraint
removal replaces $\Delta$ with a new domain $\Delta'$ and removes the $x \ne c$
constraints. The resulting node is denoted as
$\CR(\Delta' \gets \Delta \setminus \{\,c\,\})$. Finally, \emph{caching} detects
when the input formula $\phi$ is equal to a previously encountered formula
$\psi$ except for having different domains. The resulting node is denoted as
$\Reff(\sigma)$, with $\sigma$ representing the substitution mapping the domains
of $\psi$ to their corresponding domains in $\phi$.

@@

We compared \Cranetwo{}\footnote{\textsc{Crane} has two modes that decide how
  compilation rules are applied to formulas: one that uses greedy search and
  another that combines greedy and breadth-first search. In our experiments, we
  refer to these modes as \Cranegreedy{} and \Cranebfs{}, respectively.} with
\textsc{FastWFOMC} and \textsc{ForcLift} on two problems previously considered
by Dilkas and Belle~\shortcite{DBLP:conf/kr/DilkasB23}. The first problem is the
function-counting problem from \cref{example:functions}, and the second problem
is the bijection-counting problem described below. It is important to note that
comparing \Cranetwo{} and \textsc{FastWFOMC} on a more substantial set of
benchmarks is challenging because there is no automated way to translate a
formula in \FO{} or \Ctwo{} into \UFO{}, or even check if such an encoding is
possible. We ran the experiments on an AMD~Ryzen~7~5800H processor with
\SI{16}{\gibi\byte} of memory and Arch Linux~6.8.2-arch2-1 operating system.
\textsc{FastWFOMC} was executed using Python~3.8.19 with Python-FLINT~0.5.0.

@@

In this work, we have presented several contributions. First, we have developed
algorithmic techniques to find the base cases of recursive functions generated
by the original \textsc{Crane} algorithm. Second, we have extended the smoothing
procedure of \textsc{ForcLift} and \textsc{Crane} to support base case formulas
and the compilation rules introduced by Dilkas and
Belle~\shortcite{DBLP:conf/kr/DilkasB23}. Third, we have proposed an approach to
compile function definitions into C++ programs with support for
arbitrary-precision arithmetic. Lastly, we have provided experimental evidence
demonstrating that \Cranetwo{} can scale to much larger domain sizes than
\textsc{FastWFOMC} while handling more formulas than \textsc{ForcLift}.

@@

\bibliographystyle{kr}
\bibliography{paper}

>>>>

\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
\usepackage[submission]{aaai25}  % DO NOT CHANGE THIS
\usepackage{times}  % DO NOT CHANGE THIS
\usepackage{helvet}  % DO NOT CHANGE THIS
\usepackage{courier}  % DO NOT CHANGE THIS
\usepackage[hyphens]{url}  % DO NOT CHANGE THIS
\usepackage{graphicx} % DO NOT CHANGE THIS
\urlstyle{rm} % DO NOT CHANGE THIS
\def\UrlFont{\rm}  % DO NOT CHANGE THIS
\usepackage{natbib}  % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\usepackage{caption} % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\frenchspacing  % DO NOT CHANGE THIS
\setlength{\pdfpagewidth}{8.5in} % DO NOT CHANGE THIS
\setlength{\pdfpageheight}{11in} % DO NOT CHANGE THIS

\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{microtype}
\usepackage{mathtools}
\usepackage{forest}
\usepackage{siunitx}
\usepackage{amsfonts}
\usepackage{pifont}
\usepackage[inline]{enumitem}
\usepackage[capitalize,noabbrev]{cleveref}
\usepackage{booktabs}
\usepackage{amsthm}

\newtheorem{theorem}{Theorem}
\newtheorem{assumption}{Assumption}
\newtheorem{corollary}{Corollary}
\newtheorem{fact}{Observation}
\newtheorem{lemma}{Lemma}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

@@

@@

\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}
\newcommand{\expr}{\mathtt{expr}}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\FO}{$\mathsf{FO}$}
\newcommand{\UFO}{$\mathsf{UFO}^{2} + \mathsf{EQ}$}
\newcommand{\Cranetwo}{\textsc{Crane2}}
\newcommand{\Cranebfs}{\textsc{Crane2-BFS}}
\newcommand{\Cranegreedy}{\textsc{Crane2-Greedy}}

\newcommand{\crefrangeconjunction}{--}
\crefname{line}{line}{lines}
\crefname{fact}{Observation}{Observations}
\crefname{assumption}{Assumption}{Assumptions}
\crefalias{enumi}{type}
\crefname{type}{Type}{Types}
\creflabelformat{type}{#2\textup{#1}#3}
\crefalias{enumi}{step}
\crefname{step}{Step}{Steps}
\creflabelformat{step}{#2\textup{#1}#3}
\crefalias{clause}{equation}
\crefname{clause}{Clause}{Clauses}
\creflabelformat{clause}{#2\textup{(#1)}#3}
\crefalias{formula}{equation}
\crefname{formula}{Formula}{Formulas}
\creflabelformat{formula}{#2\textup{(#1)}#3}

\DeclareMathOperator{\CR}{CR}
\DeclareMathOperator{\DR}{DR}
\DeclareMathOperator{\Reff}{Ref}
\DeclareMathOperator{\Doms}{Doms}

@@

\pdfinfo{
/TemplateVersion (2025.1)
}

\setcounter{secnumdepth}{2} %May be changed to 1 or 2 if section numbers are desired.

\title{Towards Practical First-Order Model Counting}
\author{Anonymous Submission}
\affiliations{}

@@

The complexity of WFOMC is typically characterised in terms of \emph{data
  complexity}. This involves fixing the formula and determining whether an
algorithm exists that can compute the WFOMC in time polynomial with respect to
the domain size(s). If such an algorithm exists, the formula is called
\emph{liftable}~\cite{DBLP:conf/starai/JaegerB12}.
\citeauthor{DBLP:conf/pods/BeameBGS15}~\shortcite{DBLP:conf/pods/BeameBGS15}
demonstrated the existence of an unliftable formula with three variables. It is
also known that all formulas with up to two variables are
liftable~\cite{DBLP:conf/nips/Broeck11,DBLP:conf/kr/BroeckMD14}. The liftable
fragment of formulas with two variables has been expanded with various
axioms~\cite{DBLP:conf/aaai/TothK23,DBLP:journals/ai/BremenK23}, counting
quantifiers~\cite{DBLP:journals/jair/Kuzelka21} and in other
ways~\cite{DBLP:conf/nips/KazemiKBP16}.

@@

  % TODO: Yes, as mentioned at the end of Section 2.1, clauses c and d written
  % on separate lines are to be interpreted as the conjunction of c and d (maybe
  % move that sentence here?).
  Equivalently, in \FO{} we would write
  \begin{equation}\label[formula]{eq:fo}
    \begin{gathered}
      \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y)\\
      \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z.
    \end{gathered}
  \end{equation}
  The first clause asserts that each $x$ must have at least one corresponding
  $y$, while the second statement adds the condition that if $x$ is mapped to
  both $y$ and $z$, then $y$ must equal $z$. It is important to note that
  \cref{eq:fo} is written with two domains instead of just one. However, we can
  still determine the number of endofunctions by assuming that the sizes of
  $\Gamma$ and $\Delta$ are equal. This formulation, as observed by
  \citeauthor{DBLP:conf/kr/DilkasB23}~\shortcite{DBLP:conf/kr/DilkasB23}, can
  prove beneficial in enabling first-order knowledge compilation algorithms to
  find efficient solutions.
\end{example}

@@

% TODO: F missing (also not explained that it’s not needed below). E is not explained other than that we take the example from X. Would not be a problem if it is motivated differently and the focus is on post-processing (defining inputs properly) and not on the entire algorithm.
% TODO: include a high-level overview of first-order knowledge compilation and the meaning of the resulting functions. For the most part, the compilation process that constructs these function definitions can be treated as a black box for the purposes of this work. However, a few high-level details still ought to be mentioned:
% 1. Each algebraic variable ("m", "n", and "l" in the example) corresponds to a domain size.
% 2. \mathcal{D} indicates that "m" denotes the size of \Gamma, "n" denotes the size of \Delta, etc.
% 3. \mathcal{E} can contain any (positive) number of functions.
% 4. Function denoted by "f" represents the solution to the model counting problem.
% 3. The model count of the original formula is computed by evaluating f on the desired domain sizes. For example, f(3, 2) is the model count when |\Gamma| = 3, and |\Delta| = 2.
\begin{example}[Counting bijections]\label{example:overall}
  Consider the following formula (previously examined by
  \citeauthor{DBLP:conf/kr/DilkasB23}~\shortcite{DBLP:conf/kr/DilkasB23}) that
  defines predicate $P$ as a bijection between two sets $\Gamma$ and $\Delta$:
  \[
    \begin{gathered}
      \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y)\\
      \forall y \in \Delta\text{. }\exists x \in \Gamma\text{. }P(x, y)\\
      \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z\\
      \forall x, z \in \Gamma\text{. }\forall y \in \Delta\text{. }P(x, y) \land P(z, y) \Rightarrow x = z.
    \end{gathered}
  \]
  We specifically examine the first solution returned by \Cranetwo{} for this
  formula.

@@

\emph{Smoothness} is a property originating in propositional knowledge
compilation, where it refers to the situation where all disjuncts in a
disjunction node contain the same atoms~\cite{darwiche2001tractable}.
\citeauthor{DBLP:conf/ijcai/BroeckTMDR11}~\shortcite{DBLP:conf/ijcai/BroeckTMDR11}
extend this concept to first-order logic, introducing set-disjunction and
inclusion-exclusion nodes in addition to disjunction.

@@

The remainder of this section presents an extension to the smoothing algorithm
of
\citeauthor{DBLP:conf/ijcai/BroeckTMDR11}~\shortcite{DBLP:conf/ijcai/BroeckTMDR11}.
\Cref{sec:smoothingbase} explains the role of smoothing in the base-case-finding
algorithm described in \cref{sec:main}. \Cref{sec:smoothingfcg1} demonstrates
how to adapt smoothing to the compilation rules introduced by
\citeauthor{DBLP:conf/kr/DilkasB23}~\shortcite{DBLP:conf/kr/DilkasB23}.

@@

Before discussing the proposed changes to smoothing, let us briefly review the
compilation rules and their corresponding node types introduced by
\citeauthor{DBLP:conf/kr/DilkasB23}~\shortcite{DBLP:conf/kr/DilkasB23}.
\emph{Domain recursion} involves selecting a domain $\Delta$, introducing a new
constant $c \in \Delta$, and modifying the formula based on two possibilities
for each variable $x$ quantified over $\Delta$: $x = c$ or $x \ne c$. The
resulting node is denoted as $\DR(c \in \Delta)$. \emph{Constraint removal}
applies to formulas where each variable $x$ quantified over a domain $\Delta$ is
followed by the inequality constraint $x \ne c$. These formulas can be rewritten
so that each clause begins with $\forall x \in \Delta\text{.
}x \ne c \Rightarrow \ldots$ Constraint removal replaces $\Delta$ with a new
domain $\Delta'$ and removes the $x \ne c$ constraints. The resulting node is
denoted as $\CR(\Delta' \gets \Delta \setminus \{\,c\,\})$. Finally,
\emph{caching} detects when the input formula $\phi$ is equal to a previously
encountered formula $\psi$ except for having different domains. The resulting
node is denoted as $\Reff(\sigma)$, with $\sigma$ representing the substitution
mapping the domains of $\psi$ to their corresponding domains in $\phi$.

@@

We compared \Cranetwo{}\footnote{\textsc{Crane} has two modes that decide how
  compilation rules are applied to formulas: one that uses greedy search and
  another that combines greedy and breadth-first search. In our experiments, we
  refer to these modes as \Cranegreedy{} and \Cranebfs{}, respectively.} with
\textsc{FastWFOMC} and \textsc{ForcLift} on two problems previously considered
by \citeauthor{DBLP:conf/kr/DilkasB23}~\shortcite{DBLP:conf/kr/DilkasB23}. The
first problem is the function-counting problem from \cref{example:functions},
and the second problem is the bijection-counting problem described below. It is
important to note that comparing \Cranetwo{} and \textsc{FastWFOMC} on a more
substantial set of benchmarks is challenging because there is no automated way
to translate a formula in \FO{} or \Ctwo{} into \UFO{}, or even check if such an
encoding is possible. We ran the experiments on an AMD~Ryzen~7~5800H processor
with \SI{16}{\gibi\byte} of memory and Arch Linux~6.8.2-arch2-1 operating
system. \textsc{FastWFOMC} was executed using Python~3.8.19 with
Python-FLINT~0.5.0.

@@

In this work, we have presented several contributions. First, we have developed
algorithmic techniques to find the base cases of recursive functions generated
by the original \textsc{Crane} algorithm. Second, we have extended the smoothing
procedure of \textsc{ForcLift} and \textsc{Crane} to support base case formulas
and the compilation rules introduced by
\citeauthor{DBLP:conf/kr/DilkasB23}~\shortcite{DBLP:conf/kr/DilkasB23}. Third,
we have proposed an approach to compile function definitions into C++ programs
with support for arbitrary-precision arithmetic. Lastly, we have provided
experimental evidence demonstrating that \Cranetwo{} can scale to much larger
domain sizes than \textsc{FastWFOMC} while handling more formulas than
\textsc{ForcLift}.

@@

\bibliography{paper}
